CODECHECK: An open-science initiative for the independent execution of computations underlying research articles

### 2021-01-12 codecheck-paper.tex
update workflow description (clarifycations)

CODECHECK is best demonstrated by way of our example workflow, and later
we expand on the underlying principles. The workflow involves three
groups of people: (1) the AUTHOR of a paper providing the code to be checked, (2)
the PUBLISHER of a journal interested in publishing the paper by the
AUTHOR and (3) the CODECHECKER who checks that the AUTHOR's code works.
The workflow that we have refined is documented in Figure~\ref{fig:workflow}.

% TODO: how is a CODECHECK different from a "specialist reviewer" ? Is it because it is not special, but regular?

@@

\textbf{Step 1:} the AUTHOR submit their manuscript along with code and data to
the PUBLISHER. The code and data need not be openly available at this point.
However, in many cases the code and data may be published on a code hosting platform,
such as GitHub or GitLab. Ideally, the AUTHOR is expecting the CODECHECK and
prepares for it, e.g., by asking a colleague to briefly attempt a reproduction
of the submitted workflow.

\textbf{Step 2:} the publisher finds a CODECHECKER to check the code. This is
analogous to the publisher finding one or more reviewers to evaluating
the paper. However, a crucial difference is that we suggest the CODECHECKER
and the author can talk freely to each other directly (see next).

\textbf{Step 3:} the CODECHECKER runs the code, based on instructions provided by
the author, and checks if some/all of the results from the paper can be
reproduced. (See later for exact nature of ``reproduced''). If there are
any problems running the code, the CODECHECKER asks the AUTHOR for help
to resolve the problems. The burden to provide reproducible material lies with
the author.
The CODECHECKER then tries to run the code again.
This process iterates until either the CODECHECKER is successful,
or the CODECHECKER concludes the code does not work (\emph{TODO NOT
  YET HANDLED}). As part of this process, the CODECHECKER could work entirely
locally on their own compute resource, or in the cloud, e.g.~using the open
MyBinder infrastructure \cite{jupyter_binder_2018} or alternatives, some of which
are more tailored for scientific publications while others offer commercial
options to, e.g., publishers (cf. \cite{konkol_publishing_2020}).
Using such cloud-based 
infrastructure allows for the CODECHECKER and author to collaboratively improve
the code and enforces a much more complete definition of the computing environment,
but, unless secure infrastructure is provided by, e.g., by the publisher, requires code
and data to be published openly online.
Note that the task of the CODECHECKER is to check the ``mechanics'' of the
workflow, not to evaluate its correctness. 
Stodden et~al. \cite{stodden_setting_2013} distinguish between verification and
validation in the context of Mathematics, and in the same vein a CODECHECK
ensures a verification\footnote{Note this is a different use as the \emph{Verification Reports} article type in the journal \emph{Cortex}
\cite{chambers_verification_2020}},
i.e., checking that the code correctly produces the 
output it claims to create, but not a validation, i.e., does the code implement
the right algorithm to solve the problem that is researched, 
of the computational results.
Nevertheless, the simple attempt to reproduce may highlight shortcomings in
a submission with respect to transparency as it may be required by a journal's
policy and help to resolve misunderstandings and discrepancies
(cf.~\cite{christian_journal_2020}).

\textbf{Step 4:} the CODECHECKER writes a certificate stating how the code was
run and includes a copy of outputs (figures, tables) that were
independently generated. The certificate may include further recommendations to
the author how to improve the material.
The plain text in the certificate is the most flexible and feasible way to 
make clear what was really checked. Readers will have very different 
understandings of what a ``green'' checkmark (cf. Step~5) means.

\textbf{Step 5:} the certificate as well as auxiliary files created
during the check, e.g., a specification of a computing environment,
helper scripts, and the code and data, if available under open
licenses, get deposited in an open archive (currently Zenodo); the
certificate is given to the PUBLISHER.

\textbf{Step 6:} the PUBLISHER can then use the certificate as part of their
supplementary records for a paper or provide it to scientific reviewers; the
PUBLISHER also gives credit to the CODECHECKER for their work by depositing the
activity appropriately in scholarly publication databases, such as ORCID.
The PUBLISHER also ensures proper publication metadata, e.g., links from the 
certificate repository to the published paper or the original code repository.
A badge or other visual aid may be applied to highlight the fact of a completed
CODECHECK and provide a incentive and reward for the, today often considered
``additional'' efforts of providing readily computationally reproducible workflows,
despite the simplification into a binary value and the possible confusion of
the meaning, especially across disciplines and methods.

>>>>

CODECHECK is best demonstrated by way of our example workflow, and later
we expand on the underlying principles. The workflow involves three
groups of people:
(1) the AUTHOR of a paper providing the code to be checked,
(2) the PUBLISHER of a journal interested in publishing the paper 
by the author, and
(3) the CODECHECKER who checks that the authors's code works.
The workflow that we have refined is documented in Figure~\ref{fig:workflow}
and consists of six steps.

@@

\textbf{Step 1:} the author submits their manuscript along with code and data to
the PUBLISHER. The code and data need not be openly available at this point.
However, in many cases the code and data may be published on a code hosting platform,
such as GitHub or GitLab. Ideally, the AUTHOR is expecting the CODECHECK and
prepares for it, e.g.,~by asking a colleague to attempt a reproduction or by 
following the PUBLISHER's guidelines for sharing computational workflows.

\textbf{Step 2:} the publisher finds a CODECHECKER to check the code. This is
analogous to the publisher finding one or more reviewers to evaluate
the paper. However, a crucial difference is that we suggest the CODECHECKER
and the author can talk freely to each other directly (see next), independent
of the openness of the scientific peer review process.

\textbf{Step 3:} the CODECHECKER runs the code, based on instructions provided by
the author. They check if some or all of the results from the paper can be
reproduced. If there are
any problems running the code, the CODECHECKER asks the AUTHOR for help,
updates, or further documentation.
The burden to provide reproducible material lies with the author.
The CODECHECKER then tries to run the code again.
This process iterates until either the CODECHECKER is successful,
or the CODECHECKER concludes the code does not work.
As part of this process, the CODECHECKER could work entirely
locally on their own compute resource, or in the cloud, e.g.,~using the open
MyBinder infrastructure \cite{jupyter_binder_2018} or alternatives, some of which
are more tailored for scientific publications while others offer commercial
options to, e.g.,~publishers (cf. \cite{konkol_publishing_2020}).
Cloud-based 
infrastructure allows for the CODECHECKER and author to collaboratively improve
the code and enforces a complete definition of the computing environment,
but, unless secure infrastructure is provided, e.g.,~by the PUBLISHER,
requires code and data to be published openly online.

Note that the task of the CODECHECKER is to check the ``mechanics'' of the
workflow, not to evaluate its correctness. 
Stodden et~al. \cite{stodden_setting_2013} distinguish between \emph{verification}
and \emph{validation}
in the context of Mathematics. Following their definition, a CODECHECK
ensures a verification\footnote{Note this is a different use as the
\emph{Verification Reports} 
article type in the journal \emph{Cortex} \cite{chambers_verification_2020}},
of computational results, i.e.,~checking that the code correctly produces the 
output it claims to create, but not a validation, i.e.,~does the code implement
the right algorithm to solve the problem that is researched.
Nevertheless, the simple attempt to reproduce may highlight shortcomings in
a submission with respect to transparency as it may be required by a journal's
policy and help to resolve misunderstandings and discrepancies
(cf.~\cite{christian_journal_2020}).

\textbf{Step 4:} the CODECHECKER writes a certificate stating how the code was
run and includes a copy of outputs (figures, tables) that were
independently generated. The certificate may include further recommendations to
the author how to improve the material.
The free text in the certificate is the most flexible and feasible way to 
make clear what was really checked, because each workflow is unique.
Since no specific tool or platform should be required so that no authors are
excluded, automation or fixed checklists are futile.

\textbf{Step 5:} the certificate as well as auxiliary files created
during the check, e.g.,~a specification of a computing environment, data 
subsets or helper scripts, and the original code and data get deposited in
an open archive unless restrictions regarding data size, license or 
sensitivity apply.
Currently, the CODECHECKERs deposit the material on Zenodo themselves, 
while processes managed by the PUBLISHER may have them complete this step.
A badge or other visual aid may be added to the deposit and the paper and
link to the certificate.
Despite the simplification into a binary value and the risk of confusion of
the extent of the check, especially across disciplines and methods,
a badge can provide recognition value and creditation for the completed
CODECHECK. The badge and the actual check are incentives for the efforts
of providing a readily computationally reproducible workflows.

\textbf{Step 6:} the PUBLISHER can, depending on the timing, provide the
certificate to scientific reviewers or handling editors, or publish it
and link between the certificate, the paper, and any repositories 
through metadata. Currently, the CODECHECKER creates these on Zenodo\footnote{
See ``related identifiers'' and ``alternate identifiers'' in the right-hand
sidebar of certificate \texttt{2020-025} \cite{cert-2020-025}.}.
The PUBLISHER also gives credit to the CODECHECKER for their work by depositing
the activity appropriately in scholarly profiles, such as ORCID\footnote{
See peer review contributions in ORCID records at
\url{https://support.orcid.org/hc/en-us/articles/360006971333-Peer-Review}.},
The PUBLISHER also ensures proper publication metadata, e.g.,~links from the 
certificate repository to the published paper or the original code repository.
